Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-07-15 jso.qmd
proofread

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. The optimisation in PP can be non-trivial, involving non-smooth functions and optima with a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated along with JSO hyper-parameters to determine their effects on JSO success rate. Our numerical results confirm the positive impact of these metrics on the JSO success rate, with squintability being the most significant. The JSO algorithm has been implemented in the `tourr` package and functions to calculate smoothness and squintability of indexes are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
---

@@

Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising a objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data $X \in \mathbf{R}^{n\times p}$ and an index
function $f(\cdot)$, PP finds the orthonormal basis
$A \in \mathbf{R}^{p \times d}$ that maximises the index value of the
projection, $Y = XA$:

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieving a globally optimal viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP, improving optimisation (e.g. @posse1995, @marie-sainte2010,
@grochowski2011) and extending its use to visually explore
high-dimensional data. @cook1995grand introduced the PP guided tour,
which monitors the optimisation visually to see the data projections
leading in and out of the optima. An implementation is available in the
R [@R] package `tourr` [@tourr]. @RJ-2021-105 illustrated how to
diagnose optimisation processes, particularly focusing on the guided
tour, and revealed a need for improved optimisation. While improving
the quality of the optimisation solutions in the tour is essential, it
is also important to be able to view the data projections as the
optimisation progresses. Integrating the guided tour with a global
optimisation algorithm that is efficient in finding the global optimal
and enables viewing of the projected data during the exploration process
is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) to be
integrated with the projection pursuit guided tour is explored. JSO
[@chou_novel_2021] -[@rajwar_exhaustive_2023], inspired by the search
behaviour of jellyfish in the ocean, is a swarm-based metaheuristic
designed to solve global optimisation problems. Compared to traditional
methods, JSO demonstrated stronger search ability and faster
convergence, and requires fewer tuning parameters. These practical
advantages make JSO a promising candidate for enhancing PP optimisation.
<!--The study reported here explores JSO's potential in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.-->

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness as introduced in @laa_using_2020. Here we
mathematically define metrics for squintability and smoothness, which is
new for the field. A series of simulation experiments using specially
structured datasets and various PP indexes are conducted to assess the
behaviour of JSO under different settings and its sensitivity to
hyper-parameter choices (number of jellyfishes and maximum number of tries).
The relationship between the JSO performance, hyper-parameter tuning and
properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes. @sec-sim-deets
details two simulation experiments to assess JSO's performance: one
comparing JSO's performance improvements relative to an existing
optimiser, Creeping Random Search (CRS), and the other studying the
impact of different PP index properties on optimisation performance.
@sec-sim-res presents the results. @sec-conclusion summarises the work
and provides suggestions for future directions.

@@

-   `search_geodesic()`: provides a pseudo-derivative optimisation. It
    searches locally for the best direction, based on differencing the
    index values for very close projections. Then it follows the
    direction along the geodesic path between planes, stopping when the
    next index value fails to increase.
-   `search_better()`: is a brute-force optimisation searching randomly
    for projections with higher index values.
-   `search_better_random()`: is essentially simulated annealing
    [@Bertsimas93] where the search space is reduced as the optimisation
    progresses.
-   `search_posse()`: implements the algorithm described in @posse95.
-   `search_polish()`: is a very localised search, to take tiny steps to
    get closer to the local maximum.

There are several PP index functions available: `holes()` and `cmass()`
[@cook1993projection]; `lda_pp()` [@lee2005projection]; `pda_pp()`
[@lee2010projection]; `dcor2d()` and `splines2d()` [@Grimm2016];
`norm_bin()` and `norm_kol()` [@huber85]; `slice_index()`
[@Laa:2020wkm]. Most are relatively simply defined, for any projection
dimension, and implemented because they are relatively easy to optimise.
A goal is to be able to incorporate more complex PP indexes, for example
based on scagnostics (@scag, @WW08).

@@

JSO mimics the natural movements of jellyfish, which include passive and
active motions driven by ocean currents and their swimming patterns,
respectively. In the context of optimization, these movements are
abstracted to explore the search space in a way that balances
exploration (searching new areas) and exploitation (focusing on
promising areas). The algorithm aims to find the optimal solution by
adapting the behaviour of jellyfish to navigate towards the best
solution over iterations [@chou_novel_2021].

To solve the optimisation problem embedded in the PP guided tour, a
starting projection, an index function and the maximum number of
iterations are provided as input. Then, the current projection is
evaluated by the index function. The projection is then moved in a
direction determined by a random factor, influenced by how far along we
are in the optimization process. Occasionally, completely new directions
may be taken like a jellyfish might with ocean currents. A new
projection is accepted if it is an improvement compared to the current
one, rejected otherwise. This process continues and iteratively improves
the projection, until the pre-specified maximum number of trials is
reached.

@@

**Output**: `optimized_projection`

**Initialize** `current_best` as the projection with the best index
value from `current_projections`, and `current_idx` as the array of
index function values for each projection in `current_projections`

**for** each trial_id in 1 to max_tries **do**

> Calculate $c_t$ based on the current_idx and max_trial

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>
> > Define trend based on the current_best and current_projections
>
> > Update each projection towards the trend using a random factor and
> > orthonormalisation
>
> **else**
>
> > **if** a random number is greater than $1 - c_t$ **then**
> >
> > > Slightly adjust each projection with a small random factor
> > > (passive)
> >
> > **else**
> >
> > > For each projection, compare with a random jelly and adjust
> > > towards or away from it (active)
>
> Update the orientation of each projection to maintain consistency
>
> Evaluate the new projections using the index function

> **if** any new projection is worse than the current, revert to the
> `current_projections` for that case
>
> > Determine the projection with the best index value as the new
> > current_best

> **if** trial_id $\geq$ max_trial, print the last best projection
> **exit**

**return** the set of projections with the updated current_best as the
optimized_projection
:::

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters that
are most relevant to our application: the number of jellyfish and the
maximum number of trials.

@laa_using_2020 has proposed five criteria for assessing projection
pursuit indexes (smoothness, squintability, flexibility, rotation
invariance, and speed). Since not all the properties affects the
execution of the optimisation, the focus here is on the first two
properties, namely *smoothness* (@sec-smoothness) and *squintability*
(@sec-squintability), and propose two metrics to evaluate them.

@@

```{=html}
<!--
An intuitive way to measure smoothness of a function would be to count
how many continuous derivatives exist. 
To help define smoothness in our
context, we can make use of the Sobolev spaces: functions $f$ in the
Sobolev space $W^{p,\infty}$ have all derivatives of order less than $p$
continuous. Smoothness would then be the highest $p$ such that the index
function belongs to $W^{p,\infty}$. We can relax this and consider
$W^{p,q}$ Sobolev spaces.

Consider the following definition of partial derivative. Let $U$ be an
open subset of $\mathbf{R}^n$ and $f: U \rightarrow \mathbf{R}$. The
partial derivative of function $f$ at the point
$\mathbf{a} = (a_1, \ldots, a_n) \in U$ with respective to the $i$-th
variable $x_i$ is defined as

$$
\frac{\delta}{\delta x_i} f(\mathbf{a}) =  \lim_{h \rightarrow 0} \frac{f(a_1, \ldots, a_{i-1}, a_i +h, a_{i+1}, \ldots, a_n) - f(a_1 , \ldots, a_i, \ldots, a_n)}{h} = \lim_{h\rightarrow 0} \frac{f(\mathbf{a} + h\mathbf{e_i}) - f(\mathbf{a})}{h}.
$$ where $\mathbf{e_i}$ is the unit vector of the $i$-th variable $x_i$.

If the derivative is not well-defined, we propose to approximate the
derivative using the following expression:

$$\frac{1}{n_i}\sum_{i} \frac{\|f(\mathbf{a} + h_i\mathbf{e_i}) - f(\mathbf{a})\|^p}{h_i}$$

where $p \in [0,1]$. The choice of $p$ reflects the penalty behaviour.
For the application in this paper, we choose $p = 1$.

-   $h_i$ is an neighbourhood area around $a$. We can insert different
    values of $h$ to approximate the above quantity and observe how this
    quantity changes.
-->
```
This subsection proposes a metric for the smoothness of a projection
pursuit index.

If a PP index function is evaluated at some random points, as is done at
the initialization stage of the jellyfish search optimization, then
these random index values can be interpreted as a random field, indexed
by a space parameter, namely the random projection angle. This analogy
suggests to use this random training sample to fit a spatial model, a
simple one being a (spatial) Gaussian process.

The distribution of a Gaussian process is fully determined by its mean
function and covariance function. The smoothness property comes into
play in the definition of the covariance function: if a PP index is very
smooth, then two close projection angles should produce close index
values (strong correlation); by contrast, if a PP index is not very
smooth, then two close projection angles might give very different index
values (fast decay of correlations with respect to distance between
angles).

@@

where $\nu>0$ is the smoothness parameter and where $\mathcal{K}_\nu$ is
the modified Bessel function. The Matérn covariance function can be
expressed analytically when $\nu$ is a half-integer, the most popular
values in the literature being $1/2$, $3/2$ and $5/2$. The parameter
$\nu$, called *smoothness parameter*, controls the decay of the
covariance function. As such, it is an appropriate measure of smoothness
of a random field.

In this context, the parameter $\nu$ is suggested as a measure of
the smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as GpGp
[@guinness2021gpgp] or ExaGeoStatR [@abdulah2023large], to fit the
hyperparameters of a GP covariance function on data. In this project, we
make use of the GpGp package.

@@

In the existing literature, it is commonly understood that a large
squint angle implies that the objective function value is close to
optimal even when the perfect view to see the structure is far away.
A small squint angle means that the PP index value improves
substantially only when the perfect view is close by. As such,
low squintability implies rapid improvement in the index value when near
the perfect view. For PP, a small squint angle is considered to be
undesirable because it means that the optimiser needs to be very close
to be able to "see" the optima. Thus, it could be difficult for the
optimiser to find the optima. The mathematical formulation
of this intuition is proposed below:

Let $A \in \mathbf{R}^{p \times 2}$ be the 2D projection basis and $A^*$
the optimal basis achieving the maximum index value for a given
projection pursuit problem defined by the shape to find, data dimension,
and the index function $f(.)$. The projection distance $d(A, A^*)$, is
defined as the Frobenius norm $\lVert . \rVert _F$, between the squared
difference of these two bases:

@@

The index value $f(XA)$ is defined as described in the introduction. We
now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

```{=html}
<!--
\[We generate random points that is beyond 1.5 projection distance and
interpolate. Then we fit a kernel or use nonlinear least squares.\]
-->
```
It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, should play an important role in the mathematical definition
of squintability. Noticing that $f$ is expected to be a decreasing
function, the squintability metric is proposed as follows:

@@

On the one hand, $-\min_{x}f'(x)$ represents the largest value of the
gradient of $-f$. On the other hand, $\underset{x}{\arg\min}f'(x)$
represents the projection distance at which this largest gradient is
attained. Based on the above discussion, it is expected that these two
values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggest that their product (@eq-squintability) should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

@@

The JSO performance is compared with the existing Creeping Random Search
(CRS) [@RJ-2021-105; @laa_using_2020] currently used in guided tour, and
explores how JSO behaves with different hyper-parameter (number of
jellyfishes and maximum tries) and data dimension combinations. A second
simulation studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the success rate of the JSO performance.

@@

The performance of JSO is investigated in comparison with the existing
optimiser, CRS, with respect to two hyper-parameters (number of
jellyfishes and maximum number of tries). The performance is measured by
success rate, defined as the proportion of simulations that achieves a
final index value within 0.05 of the best index value found among all 50
simulations. This comparison is based on projection pursuit to find the
pipe shape investigated by @laa_using_2020 using two simple and
well-behaved indexes, the `holes` index, and the `splines2d` index.
Fifty simulations are conducted with both JSO and CRS, in four different
data dimensions ($d = 6, 8, 10, 12$). JSO uses 100 jellyfishes with 100
maximum number of tries, while the CRS allows a maximum of 1000 samples
at each iteration before the algorithm terminates. The different numbers
accounts for the multiple paths of JSO to enable fairer comparison with
CRS. The results of the simulation are collected using the data
structure proposed in @RJ-2021-105 is used here for assessing JSO, where
the design parameters are stored along with index value, projection
basis, random seed, and computation time.
<!--A sample for the sine-wave structure in 6D data using a spline index function (`splines2d`) is printed below:

```{r}
dplyr::glimpse(sim_sine_6d_spline_head |> select(-id, -info)) 
```
-->

In this simulation, 50 jellyfishes (`n_jellies`) and a maximum of 50 tries
(`max_tries`) are used. The full data contains 50 simulations, indexed
by the variable `sim`, with the associated `seed`s. Recorded in columns
`basis`, `index_val`, `tries`, `loop` and `time`, are the projection
basis, index value, iteration ID, jellyfish ID, and computational time,
respectively.
<!-- The basis column records every visited basis before -->
<!-- comparing with the current (last) basis. If a visited basis has a -->
<!-- smaller index value than the current (last) one, JSO will keep the -->
<!-- current basis for the next iteration in the search but still records the -->
<!-- visited inferior basis.  -->

In the second experiment, 50 simulations are conducted at each
hyper-parameter combination to analyse its effect on the JSO success
rate: number of jellyfishes (20, 50, and 100), and maximum number of tries
(50 and 100). @fig-success-rate illustrates the calculation based on the
projection found by the above simulation. Here, the optimal index value
is
`r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`.
Out of the 50 simulations,
`r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb`
achieved an index value within \[`r aa - 0.05`, `r aa`\], resulting in a
success rate of
`r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`.

<!-- @fig-projection shows the best projection found in each simulation, -->

<!-- sorted by the index value. Most of the projections show a clear -->

<!-- sine-wave structure, while the last six show only a straight line -->

<!-- with clusters at the end. When local optima exist in the projection -->

>>>>

---
title: Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit
header-includes:
   - \usepackage{algorithm}
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. The optimisation in PP can be non-trivial, involving non-smooth functions and optima with a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, this work proposes novel methods to quantify two properties of the PP index -- smoothness and squintability -- that capture the complexities inherent in PP optimisation problems. These two metrics are evaluated along with JSO hyper-parameters to determine their effects on JSO success rate. Our numerical results confirm the positive impact of these metrics on the JSO success rate, with squintability being the most significant. The JSO algorithm has been implemented in the `tourr` package and functions to calculate smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal of Multivariate Analysis
      formatting: preprint
      model: 3p
      cite-style: numbername
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
---

@@

Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising an objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data $X \in \mathbf{R}^{n\times p}$ and an index
function $f(\cdot)$, PP finds the orthonormal basis
$A \in \mathbf{R}^{p \times d}$ that maximises the index value of the
projection, $Y = XA$:

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieve a globally optimal viewpoint of the data.

Optimisation of PP is typically investigated in the literature when new
indexes are proposed [@posse1995; @marie-sainte2010; @grochowski2011] or
when visualisation method are used to track the optimisation process.
@cook1995grand introduced the PP guided tour, which monitors the
optimisation visually to see the data projections leading in and out of
the optima. An implementation is available in the `tourr` package
[@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose
optimisation processes, particularly focusing on the guided tour, and
revealed a need for improved optimisation. While improving the quality
of the optimisation solutions in the tour is essential, it is also
important to be able to view the data projections as the optimisation
progresses. Integrating the guided tour with a global optimisation
algorithm that is efficient in finding the global optimal and enables
viewing of the projected data during the exploration process is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) to be
integrated with the projection pursuit guided tour is explored. JSO,
[@chou_novel_2021] - [@rajwar_exhaustive_2023], inspired by the search
behaviour of jellyfish in the ocean, is a swarm-based metaheuristic
designed to solve global optimisation problems. Compared to traditional
methods, JSO has demonstrated stronger search ability and faster
convergence, and requires fewer tuning parameters. These practical
advantages make JSO a promising candidate for enhancing PP optimisation.
<!--The study reported here explores JSO's potential in enhancing projection pursuit and examines the characteristics
and advantages of such implementation.-->

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness as introduced in @laa_using_2020. Here we
mathematically define metrics for squintability and smoothness, which is
new for the field. A series of simulation experiments using various
datasets and PP indexes are conducted to assess JSO's behaviour and its
sensitivity to hyper-parameter choices (number of jellyfish and maximum
number of tries). The relationship between the JSO performance,
hyper-parameter choices and properties of PP indexes (smoothness and
squintability) is analysed to provide guidance on selecting optimisers
for practitioners using projection pursuit. Additionally, it aims to
guide the design of new indexes that facilitate easy optimization for PP
researchers.

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes, smoothness and
squintability. @sec-sim-deets details two simulation experiments to
assess JSO's performance: one comparing JSO's performance improvements
relative to an existing optimiser, Creeping Random Search (CRS), and the
other studying the impact of different PP index properties on
optimisation performance. @sec-sim-res presents the results.
@sec-conclusion summarises the work and provides suggestions for future
directions.

@@

-   `search_geodesic()`: provides a pseudo-derivative optimisation. It
    searches locally for the best direction, based on differencing the
    index values for very close projections. Then it follows the
    direction along the geodesic path between planes, stopping when the
    next index value fails to increase.
-   `search_better()`: also known as Creeping Random Search (CRS), is a
    brute-force optimisation searching randomly for projections with
    higher index values.
-   `search_better_random()`: is essentially simulated annealing
    [@Bertsimas93] where the search space is reduced as the optimisation
    progresses.
-   `search_posse()`: implements the algorithm described in @posse95.
-   `search_polish()`: is a very localised search, to take tiny steps to
    get closer to the local maximum.

There are several PP index functions available: `holes()` and `cmass()`
[@cook1993projection]; `lda_pp()` [@lee2005projection]; `pda_pp()`
[@lee2010projection]; `dcor2d()` and `splines2d()` [@Grimm2016];
`norm_bin()` and `norm_kol()` [@huber85]; `slice_index()`
[@Laa:2020wkm]. Most are relatively simply defined, for any projection
dimension, and implemented because they are relatively easy to optimise.
A goal is to be able to incorporate more complex PP indexes, for
example, based on scagnostics (@scag, @WW08).

@@

JSO mimics the natural movements of jellyfish, which include passive and
active motions driven by ocean currents and their swimming patterns,
respectively. In the context of optimization, these movements are
abstracted to explore the search space, aiming to balance exploration
(searching new areas) and exploitation (focusing on promising areas).
The algorithm aims to find the optimal solution by adapting the
behaviour of jellyfish to navigate towards the best solution over
iterations [@chou_novel_2021].

To solve the optimisation problem embedded in the PP guided tour, a
starting projection, an index function, the number of jellyfish, and the
maximum number of trials (tries) are provided as input. Then, the
current projection is evaluated by the index function. The projection is
then moved in a direction determined by a random factor, influenced by
how far along we are in the optimisation process. Occasionally,
completely new directions may be taken like a jellyfish might with ocean
currents. A new projection is accepted if it is an improvement compared
to the current one, rejected otherwise. This process continues and
iteratively improves the projection, until the pre-specified maximum
number of trials is reached.

@@

**Output**: `optimised_projection`

**Initialize** `current_best` as the projection with the best index
value from `current_projections`, and `current_idx` as the array of
index values for each projection in `current_projections`

**for** each `trial_id` in 1 to `max_tries` **do**

> Calculate the time control value, $c_t$, based on `current_idx` and
> `max_trial`

> **if** $c_t$ is greater than or equal to $0.5$ **then**
>
> > Define trend based on the `current_best` and `current_projections`
>
> > Update each projection towards the trend using a random factor and
> > orthonormalisation
>
> **else**
>
> > **if** a random number is greater than $1 - c_t$ **then**
> >
> > > Slightly adjust each projection with a small random factor
> > > (passive)
> >
> > **else**
> >
> > > For each projection, compare with a random jellyfish and adjust
> > > towards or away from it (active)
>
> Update the orientation of each projection to maintain consistency
>
> Evaluate the new projections using the index function

> **if** any new projection is worse than the current, revert to the
> `current_projections` for that case
>
> > Determine the projection with the best index value as the new
> > `current_best`

> **if** `trial_id` $\ge$ `max_trial`, print the last best projection
> **exit**

**return** the set of projections with the updated `current_best` as the
`optimised_projection`
:::

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of tries.

@laa_using_2020 has proposed five criteria for assessing projection
pursuit indexes (smoothness, squintability, flexibility, rotation
invariance, and speed). Since not all index properties affect the
optimisation process, the focus here is on the first two properties,
*smoothness* (@sec-smoothness) and *squintability* (@sec-squintability),
for which metrics are proposed to quantify them.

@@

This subsection proposes a metric for the smoothness of a projection
pursuit index.

If a PP index function is evaluated at some random points, as is done at
the initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process.

The distribution of a Gaussian process is fully determined by its mean
and covariance function. The smoothness property comes into play in the
definition of the covariance function: if a PP index is very smooth,
then two close projection angles should produce close index values
(strong correlation); by contrast, if a PP index is not very smooth,
then two close projection angles might give very different index values
(fast decay of correlations with respect to distance between angles).

@@

where $\nu>0$ is the smoothness parameter and where $\mathcal{K}_\nu$ is
the modified Bessel function. The Matérn covariance function can be
expressed analytically when $\nu$ is a half-integer, the most popular
values in the literature being $\frac{1}{2}$, $\frac{3}{2}$ and
$\frac{5}{2}$. The parameter $\nu$, called *smoothness parameter*,
controls the decay of the covariance function. As such, it is an
appropriate measure of smoothness of a random field.

In this context, the parameter $\nu$ is suggested as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as `GpGp`
[@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large], to fit the
hyperparameters of a GP covariance function on data. In this project,
the `GpGp` package is used.

@@

From @barnett1981interpreting and @laa_using_2020, a large squint angle
implies that the objective function value is close to optimal even when
the perfect view to see the structure is far away. A small squint angle
means that the PP index value improves substantially only when the
perfect view is close by. As such, low squintability implies rapid
improvement in the index value when near the perfect view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimiser needs to be very close to be able to "see" the optima.
Thus, it could be difficult for the optimiser to find the optima. The
mathematical formulation of this intuition is proposed below:

Let $A \in \mathbf{R}^{p \times d}$ be the d-D projection basis and
$A^*$ be the optimal basis achieving the maximum index value for a given
projection pursuit problem defined by the shape to find, data dimension,
and the index function $f(.)$. The projection distance $d(A, A^*)$, is
defined as the Frobenius norm $\lVert . \rVert _F$, between the squared
difference of these two bases:

@@

We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.
Noticing that $f$ is expected to be a decreasing function, the
squintability metric is proposed as follows:

@@

$-\min_{x}f'(x)$ represents the largest value of the gradient of $-f$
and $\underset{x}{\arg\min}f'(x)$ represents the projection distance at
which this largest gradient is attained. It is expected that these two
values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggest that their product (@eq-squintability) should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

@@

The JSO performance is compared with an existing optimiser, Creeping
Random Search (CRS) [@RJ-2021-105; @laa_using_2020] used in the PP
guided tour to explore JSO's behaviour under different hyper-parameter
and data dimension combinations. The second simulation studies the
effect of index properties (smoothness and squintability), along with
JSO hyper-parameters, and data dimension, on the success rate of the JSO
performance. This section describes the simulation details, with the
results deferred to Section @sec-sim-res.

@@

he performance of JSO is investigated both in comparison to the existing
optimizer, CRS, and across various hyper-parameter values. The
performance is measured by the success rate, defined as the proportion
of simulations that achieves a final index value within 0.05 of the best
index value found among all 50 simulations (see @fig-success-rate for an
illustration). This comparison is based on projection pursuit to find
the pipe shape investigated by @laa_using_2020 using the `holes` index.

Fifty simulations are conducted with both JSO and CRS, in four different
data dimensions ($d = 6, 8, 10, 12$). JSO uses 100 jellyfishes with a
maximum of 100 tries, while the CRS allows a maximum of 1000 samples at
each iteration before the algorithm terminates. The different numbers
account for the multiple paths of JSO to enable fairer comparison with
CRS. The results of the simulation are collected using the data
structure proposed in @RJ-2021-105 for assessing JSO, where the design
parameters are stored along with index value, projection basis, random
seed, and computation time.

Fifty additional simulations are conducted for each hyper-parameter
combination to analyze how they affect the JSO success rate. This
includes variations in the number of jellyfish (20, 50, and 100) and the
maximum number of tries (50 and 100).

```{r}
#| label: fig-success-rate
#| fig.cap: "Illustration of success rate calculation: Final projections based on projection pursuit to find the pipe shape in 8D data using the holes index, optimised by CRS, in 50 simulations. The 50 final projections are sorted by their index values. The highest index value found across all simulations is 0.969. Out of the 50 simulations, 43 achieved an index value within 0.05 of the best, resulting in a success rate of 0.86 (43/50)."
# pipe_jellyfish_12d <- pipe_better |> 
#   filter(dim == 2) |> 
#   mutate(dim = as.numeric(dim) * 8, sim = as.numeric(sim)) 
# proj_dt <- map_dfr(pipe_jellyfish_12d |> pull(basis),
#               ~as_tibble(pipe1000_8d %*% .x), .id = "sim") |>
#   mutate(sim = as.numeric(sim)) |>
#   left_join(pipe_jellyfish_12d |> select(sim, index_val), by = "sim")
# 
# idx_val_sorted <- proj_dt |> pull(index_val) |> unique() |> sort()
# proj_dt |>
#   ggplot(aes(x = V1, y = V2)) +
#   geom_point(size = 0.5) +
#   geom_text(aes(label = round(index_val, 3)), x = 0, y = 3.7,
#             size = 8) +
#   xlim(-4, 4) + ylim(-4, 4) +
#   facet_wrap(~fct_reorder(as.factor(sim), -index_val), ncol = 10) +
#   theme_bw() +
#   theme(aspect.ratio = 1, axis.ticks = element_blank(),
#         axis.text = element_blank(), axis.title = element_blank(),
#         panel.grid = element_blank(), strip.background = element_blank(),
#         strip.text = element_blank()) 
knitr::include_graphics(here::here("figures", "success-rate.png"))
```
